How do changes in ex vary between countries?

### 2022-03-07 draft_paper_sandpit.Rmd

```{r}

# Let's start with a single example population. (There are many different population subgroups, hence the 
# benefit of being able to automate this later.) We're going to start with England & Wales, females, life 
# expectancy at birth (x == 0)

example_df <- 
  hmd_ex_selected_countries_with_synth %>% 
      filter(x == 0) %>% 
      filter(sex == "female") %>% 
      filter(code == "GBRTENW") %>% 
      filter(year >= 1979)

# Now we create the change in ex variable as that's the response variable in the model
example_df <- 
  example_df %>% 
    arrange(year) %>% 
    mutate(delta_ex = ex - lag(ex)) # %>% 
  # ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
  # geom_line() +
  # geom_hline(yintercept = 0) + 
  # stat_smooth()

# The null model is really simple: just an intercept 
model_null <- lm(delta_ex ~ 1, data = example_df)
summary(model_null)
#

# The code below shows how we can be even simpler, and force the regression model not to 
# even have an intercept (i.e. have zero predictor variables)

# model_nuller <- lm(delta_ex ~ 0, data = example_df)
# summary(model_nuller)

# And we can compare the intercept with the no intercept model as follows 
# anova(model_nuller, model_null)
# a statistically significant p value shows the intercept is definitely worth including in the 
# model. Substantively that it's wrong to assume no improvement on average in ex over time since the 1980s

# There are lots of possible alternative models, some examples of which are below:
model_alt01 <- lm(delta_ex ~ year >= 1990, data = example_df)
summary(model_alt01)

@@

# each of these models allows an adjustment to the horizontal line after a given year has been reached. 
# In the examples above some possible given years are 1990, 1995, 1997, 2008 and 2011. 
# Each of these is a candidate value for tau. But we've not looked at all possible/plausible values of this.

# Hence, it would be good to automate further to search for each possible tau and see how the fit and model 
# results change.

# We'll start to automate by building a function which creates T_param given tau, and runs and returns the model
# with this type of T_param
# (If tau is 1990, then T_param is equivalent to model_alt01 above; if tau were 1995, then it's equivalent to model_alt02, and so on. The only difference is that the new coefficient created has the same name, T_param, 
# whereas in the above it would have a different name each time, because the coefficient is being 
# created in situ, within the formula specification, rather than by pre-processing the dataframe)
run_alt_with_given_tau <- function(tau, df){
  df <- 
    df %>% 
      filter(!is.na(delta_ex)) %>% 
      mutate(
        T_param = ifelse(year < tau, FALSE, TRUE)
      )

@@

# We can now run the above function, which creates a model, for a wide range of possible tau values
# This is the sequence listed and named tau_candidate, which is every whole number from 
# 1982 to 2015

# For each of these models, we can then get a measure of model fit. We're going to use AIC, though 
# could have used other measures. With AIC (and with BIC, and RMSE etc) better fit is indicated with 
# a lower value
grid_search_df <- tibble(
  tau_candidate = 1982:2015 # column of candidate values
) %>% 
  mutate(
    model_alt = map(tau_candidate, run_alt_with_given_tau, df = example_df) # run the model with the value of tau in tau_candidate
  ) %>% 
  mutate(
    aic = map_dbl(model_alt, AIC)
  )

# Now we can see which proposed tau results in what fit value (AIC), initially just by graphing it.
grid_search_df %>% 
  ggplot(aes(tau_candidate, aic)) + 
  geom_point()

# So, in the above we found 2012 as tau minimised AIC. So, that's the best possible alternative model. 
# Let's run and label it as such
best_alt_model <- example_df %>% 
  mutate( T_param = ifelse(year < 2012, FALSE, TRUE)) %>% 
  lm(delta_ex ~ T_param, data = .)

# Now we want to create the null model to compare it against (which just has an intercept term alone)
null_model <- example_df %>% 
  lm(delta_ex ~ 1, data = .)

# And finally, we want to compare the best_alt_model against the null model using an F test
anova(null_model, best_alt_model)
# We are interested in the P value here. If it's statistically significant then the best_alt_model does better
# (Is worth its additional complexity in terms of how much it makes the model fit) than the null model,
# So... for females, from birth, in England & Wales, there's evidence that there WAS a significant change in 
# average life expectancy improvement from 2012 onwards. (i.e. two years after austerity was introduced)

# We can manually repeat the same exercise for some additional subsets of the data (population, gender, starting age combinations). The below shows the same for males in England & Wales, from birth.

@@

null_model <- example_df %>% 
  lm(delta_ex ~ 1, data = .)

anova(null_model, best_alt_model)

# This time 2013 is a slightly better candidate than 2012, hence 2013 being used in best_alt_model not 2012.

# Another example: France, males
###
example_df <- 
  hmd_ex_selected_countries_with_synth %>% 
      filter(x == 0) %>% 
      filter(sex == "male") %>% 
      filter(code == "FRATNP") %>% 
      filter(year >= 1979)

@@

summary(best_alt_model)

null_model <- example_df %>% 
  lm(delta_ex ~ 1, data = .)
summary(null_model)

@@

# Here 2015 has the lowest value, but this might be an artefact of the data period covered. 
# We can also see that 2012 is the next best candidate.
```

>>>>

```{r}
example_df <- 
  hmd_ex_selected_countries_with_synth %>% 
      filter(x == 0) %>% 
      filter(sex == "female") %>% 
      filter(code == "GBRTENW") %>% 
      filter(year >= 1979)

example_df <- 
  example_df %>% 
    arrange(year) %>% 
    mutate(delta_ex = ex - lag(ex)) # %>% 
  # ggplot(aes(x = year, y = delta_ex)) + geom_point() + 
  # geom_line() +
  # geom_hline(yintercept = 0) + 
  # stat_smooth()

model_null <- lm(delta_ex ~ 1, data = example_df)
summary(model_null)
# 
# model_nuller <- lm(delta_ex ~ 0, data = example_df)
# summary(model_nuller)
# 
# anova(model_nuller, model_null)

model_alt01 <- lm(delta_ex ~ year >= 1990, data = example_df)
summary(model_alt01)

@@

run_alt_with_given_tau <- function(tau, df){
  df <- 
    df %>% 
      filter(!is.na(delta_ex)) %>% 
      mutate(
        T_param = ifelse(year < tau, FALSE, TRUE)
      )

@@

grid_search_df <- tibble(
  tau_candidate = 1982:2015
) %>% 
  mutate(
    model_alt = map(tau_candidate, run_alt_with_given_tau, df = example_df)
  ) %>% 
  mutate(
    bic = map_dbl(model_alt, BIC)
  )

grid_search_df %>% 
  ggplot(aes(tau_candidate, bic)) + 
  geom_point()

null_alt_model <- example_df %>% 
  lm(delta_ex ~ 1, data = .)

@@

null_alt_model <- example_df %>% 
  lm(delta_ex ~ 1, data = .)

###
example_df <- 
  hmd_ex_selected_countries_with_synth %>% 
      filter(x == 0) %>% 
      filter(sex == "male") %>% 
      filter(code == "FRATNP") %>% 
      filter(year >= 1979)

@@

null_alt_model <- example_df %>% 
  lm(delta_ex ~ 1, data = .)

@@
